Chemical composition analysis for X-ray transport container scans. \\ Анализ химического состава при сканировании транспортных контейнеров гамма-излучением

### 2018-10-30 EMIN-2018/article/emin2108.tex
save EMIN

\newpage
\label{sec:intro}
\section*{Introduction}
Geant4~\cite{ALLISON2016186, spirin}
\section*{Dual energy method}
Consider how the flux of gamma rays decreases. Прозрачность (transmittance) описывается следующей формулой:
\begin{equation}
\label{eq:trans}
T(E_0, t, Z) = \frac{\int \limits_0^{E_0} S(E_0, E) \exp(-\mu(E,Z)\times t)~dE)}{\int \limits_0^{E_0} S(E_0, E)~dE},
\end{equation}
where $T$ -  transmittance, $S(E_0, E)$ - response function, $\mu(E,Z)$ - attenuation, $t$ -  optical thickness of material, $E_0$ -  up-limit energy of bremsstrahlung, $E$ - energy of gamma ray, $Z$ - charge of nuclei.
We assume that bremsstrahlung is used as a source of gamma rays, with a spectrum, for example, as in figure~\ref{fig01}b and with the maximum energy depending on the energy of the electron beam. Our transmittance also depends on the mean material attenuation coefficient. Figure~\ref{fig01}a shows the dependence of the attenuation coefficient on energy for different materials. We can distinguish three areas: the initial one in which the photoelectric effect dominates and only materials with a large nuclear charge stand out; the average in which Compton scattering dominates and materials are not distinguishable, and the area where the main influence is produced by the production of electron-positron pairs, and the materials are quite well distinguishable. The last area can be used for the dual energy method.
\begin{figure}[t]
\begin{center}
    % Сделать подписи к отдельным картинкам
\includegraphics[width=60mm]{figures/Attenuation.pdf} 
\includegraphics[width=60mm]{figures/Bremsstrahlung.pdf}  
\vspace{-3mm}
\caption{a)Attenuation curve b) Bremsstrahlung spectrum by $10 MeV$ electron}
\end{center}
\labelf{fig01}
\vspace{-5mm}
\end{figure}
The dual energy method is based on using two electron beams with different energies. Since the equation from the previous slide does not determine the nuclear charge, if the optical thickness of the material is not known, to solve this problem, we consider the transmittance for the two limit energies of gamma rays $E^{(1)}_0$ and $E^{(2)}_0$, and then minimizing this functional:
\begin{equation}
F(z) = \frac{|t(E^{(1)}_0,z) - t(E^{(2)}_0,z)|}{t(E^{(1)}_0,z)} \to min
\end{equation}
we exclude unknown optical thickness from consideration and determine the average charge of the material. This method allows to determine which of these four groups the scanned object belongs to. This technique allows to determine scan object as a one from four possible groups: $Z_{eff} \sim 5$, $Z_{eff} \sim 13$, $Z_{eff} \sim 26$, $Z_{eff} \sim 82$.\\
This method has several disadvantages:
    \begin{itemize}
        \item It is too difficult to irradiate the target with beams with different energy.
        \item Low efficiency for target which contains elements with strongly different charges.
    \end{itemize}
Therefore, we offer an alternative method:
    \begin{itemize}
        \item Use only one electron beam with energy $E = 10 MeV$.
        \item Measure not only the space distribution, but also the energy of gamma rays.
    \end{itemize}

\section*{Simulation}
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        \includegraphics[width=120mm]{figures/yed_schema_1.pdf}
        \vspace{-3mm}
        \caption{a) }
    \end{center}
    \labelf{pic:schema1}
    \vspace{-5mm}
\end{figure}
For evaluation, we conducted several simulations using the scheme (fig.~\ref{pic:schema1}):
An electron beam with the energy of 10 MeV collides with a wolfram converter, producing a bremsstrahlung that irradiates a steel two-meter container, inside which the object under study is located and is detected by a detector. Расcтояние между вольфрамовым конвертером и контейнером два метра, между контернером и детектором 10 см. Приведем некоторые примеры моделирования.
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        % Переделать картинку от Туринге
        \includegraphics[width=60mm]{figures/Sword.pdf} 
        \includegraphics[width=60mm]{figures/UranCube1.pdf}  
        \vspace{-3mm}
        \caption{a) b)}
    \end{center}
    \labelf{pic:sword}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        % Переделать картинку от Туринге
        \includegraphics[width=60mm]{figures/UranCube2.pdf} 
        \includegraphics[width=60mm]{figures/Hex.pdf}  
        \vspace{-3mm}
        \caption{a) b)}
    \end{center}
    \labelf{pic:hex}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        % Переделать картинку от Туринге
        \includegraphics[width=60mm]{figures/diffmat0.pdf} 
        \includegraphics[width=60mm]{figures/diffmat.pdf}  
        \vspace{-3mm}
        \caption{a) b)}
    \end{center}
    \labelf{fig01}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        % Переделать картинку от Туринге
        \includegraphics[width=60mm]{figures/diffmat1.pdf} 
        \includegraphics[width=60mm]{figures/Difference.pdf}  
        \vspace{-3mm}
        \caption{a) b)}
    \end{center}
    \labelf{pic:diff}
    \vspace{-5mm}
\end{figure}
На рисунке ~\ref{pic:sword})a пример опасного предмета из стали неравномерной толщины близкой к толщине стенок контейнера. На рисунках~\ref{pic:sword}b и ~\ref{pic:hex}a  показан результат  моделирования кубика урана с ребром 6 сантиметров (масса около 4 кг), помещённого в свинцовую сферу толщиной 1 см, как показало моделирования такой куб можно обнаружить при толщине оболочки вплоть до 5 сантиметров. Рисунок ~\ref{pic:hex}b демонстрирует различие между двумя органическими материалами: безопасным - целлюлозы, и опасным  - RDX. Различие значимо, а значит есть возможность создать алгоритмы поиска органических взрывчатых веществ. На рисуноке ~\ref{pic:diff}b  показывается результат сравнения двух энергетических спектров (в  качестве метрики выбран логарифм отношения интенсивностей) от алюминиевого и уранового шариков диаметром 1 см. Как мы видим даже на таких малых масштабах и небольших (по сравнение с реальным пучком) интенсивностях возможна регистрация различий в энергетических спектрах.

\section*{Thickness reconstruction}
\begin{wrapfigure}[9]{r}{0.6\linewidth} 
    % Сделать подписи к отдельным картинкам
    \includegraphics[width=\linewidth]{figures/yed_schema_2.pdf}
    \vspace{-3mm}
    \caption{ Thickness reconstruction}
    \labelf{schema2}
    \vspace{-5mm}
\end{wrapfigure}
Мы будем рассматривать одномерный случай, когда гамма-кванты проходят скозь стопку из нескольких материалов с фиксированной полной толщиной и нам нужно восстановить толщины отдельных материалов (см. схему ~\ref{scema2}). Мы будем использовать простую физическую модель в которой ослабление задается  следующей формулой: 
\begin{equation}
\label{eq:gamma}
\frac{N(E)}{N_0(E)} = \exp(-\sum_i \Sigma^{mean}_i(E)x_i),
\end{equation}
where $x_i$ --- thickness of the $i$-layer, $\Sigma^{mean}_i$ --- mean cross-section for group of materials with близкими зарядами ядра, $N,~N_0$ --- the number of gamma.При этом мы не будем учитывать множественное рассеяние и наличие аннигиляционной линии. Мы считаем что полная толщина известна и для восстановления толщин отдельных слоев мы будем использовать метод наименьших квадратов, то есть минимизировать такую сумму:
\begin{equation}
\sum_E(\ln \frac{N(E)}{N_0(E)} + \sum_i \Sigma^{mean}_i(E)x_i))^2 \to min
\end{equation}
Далее приведен пример работы алгоритма. We consider the object which has 3 layers of aluminium, iron and lead.The figure~\ref{rec:ex}a shows a contribution of every reconstructed material in summary attenuation. The table~\ref{tab:rec} contains the results of reconstruction.As we can see from the table, its result is quite accurate. To clarify the capabilities of the algorithm, we conducted several numerical experiments.We also used aluminum, iron and lead, but now we have taken many sets of thicknesses, with a total thickness between 30 and 180 centimeters. The figure~\ref{rec:ex}b shows the variation of recovery errors for these sets. As we can see, the thickness of heavy elements is determined best of all - with an accuracy about 5\%, and worst of all for materials from the iron group, the error there reaches 30\%. However, this is too simple model. What good will it do? We used only the energy distribution, if we also add a space distribution, then we can try to reconstruct the three-dimensional structure of the cargo in the container (the 3D gamma tomography). Therefore, our simple model shows that we have the prospect of creating a truly powerful system for analyzing the contents of containers.
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        \includegraphics[width=60mm]{figures/reconstruction.pdf} 
        \includegraphics[width=60mm]{figures/relError.pdf}  
        \vspace{-3mm}
        \caption{a) b)Distribution of reconstruction errors for various numerical experiments}
    \end{center}
    \labelf{rec:ex}
    \vspace{-5mm}
\end{figure}

\begin{table}
\begin{center}
        \begin{tabular}[c]{|c|c|c|}
        \hline 
        Material & Real, cm & Reconstructed, cm \\ 
        \hline 
        Al & 20 & 19.6 \\ 
        \hline 
        Fe & 40 & 41.6 \\ 
        \hline 
        Pb & 30 & 28.7 \\ 
        \hline 
    \end{tabular}
\end{center}
\caption{Example of thickness reconstruction}
    \label{tab:rec}
\end{table}

@@

The experiment was conducted by Dr. Guber and Dr. Ivashkin.
The sum of signals from 2 photodiode\\
Noise threshold: 100 KeV\\
\begin{table}
    \begin{center} 
        \begin{tabular}[c]{|c|c|c|}

>>>>

\newpage
\label{sec:intro}
\section*{Introduction}
Geant4~\cite{ALLISON2016186, spirin}
\section*{Dual energy method}
Consider how the flux of gamma rays decreases. Прозрачность (transmittance) описывается следующей формулой:
\begin{equation}
\label{eq:trans}
T(E_0, t, Z) = \frac{\int \limits_0^{E_0} S(E_0, E) \exp(-\mu(E,Z)\times t)~dE)}{\int \limits_0^{E_0} S(E_0, E)~dE},
\end{equation}
where $T$ -  transmittance, $S(E_0, E)$ - response function, $\mu(E,Z)$ - attenuation, $t$ -  optical thickness of material, $E_0$ -  up-limit energy of bremsstrahlung, $E$ - energy of gamma ray, $Z$ - charge of nuclei.
We assume that bremsstrahlung is used as a source of gamma rays, with a spectrum, for example, as in figure~\ref{fig01}b and with the maximum energy depending on the energy of the electron beam. Our transmittance also depends on the mean material attenuation coefficient. Figure~\ref{fig01}a shows the dependence of the attenuation coefficient on energy for different materials. We can distinguish three areas: the initial one in which the photoelectric effect dominates and only materials with a large nuclear charge stand out; the average in which Compton scattering dominates and materials are not distinguishable, and the area where the main influence is produced by the production of electron-positron pairs, and the materials are quite well distinguishable. The last area can be used for the dual energy method.
\begin{figure}[t]
\begin{center}
\includegraphics[width=60mm]{figures/Attenuation.pdf} 
\includegraphics[width=60mm]{figures/Bremsstrahlung.pdf}  
\vspace{-3mm}
\caption{a)Attenuation curve b) Bremsstrahlung spectrum by $10 MeV$ electron}
\end{center}
\labelf{fig01}
\vspace{-5mm}
\end{figure}
The dual energy method is based on using two electron beams with different energies. Since the equation from the previous slide does not determine the nuclear charge, if the optical thickness of the material is not known, to solve this problem, we consider the transmittance for the two limit energies of gamma rays $E^{(1)}_0$ and $E^{(2)}_0$, and then minimizing this functional:
\begin{equation}
F(z) = \frac{|t(E^{(1)}_0,z) - t(E^{(2)}_0,z)|}{t(E^{(1)}_0,z)} \to min
\end{equation}
we exclude unknown optical thickness from consideration and determine the average charge of the material. This method allows to determine which of these four groups the scanned object belongs to. This technique allows to determine scan object as a one from four possible groups: $Z_{eff} \sim 5$, $Z_{eff} \sim 13$, $Z_{eff} \sim 26$, $Z_{eff} \sim 82$.\\
This method has several disadvantages:
    \begin{itemize}
        \item It is too difficult to irradiate the target with beams with different energy.
        \item Low efficiency for target which contains elements with strongly different charges.
    \end{itemize}
Therefore, we offer an alternative method:
    \begin{itemize}
        \item Use only one electron beam with energy $E = 10 MeV$.
        \item Measure not only the space distribution, but also the energy of gamma rays.
    \end{itemize}

\section*{Simulation}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=120mm]{figures/yed_schema_1.pdf}
        \vspace{-3mm}
        \caption{a) Схема моделирования }
    \end{center}
    \labelf{pic:schema1}
    \vspace{-5mm}
\end{figure}
For evaluation, we conducted several simulations using the scheme (fig.~\ref{pic:schema1}):
An electron beam with the energy of 10 MeV collides with a wolfram converter, producing a bremsstrahlung that irradiates a steel two-meter container, inside which the object under study is located and is detected by a detector. Расcтояние между вольфрамовым конвертером и контейнером два метра, между контернером и детектором 10 см. Приведем некоторые примеры моделирования.
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=60mm]{figures/Sword.pdf} 
        \includegraphics[width=60mm]{figures/UranCube1.pdf}  
        \vspace{-3mm}
        \caption{a)Опасный стальной объект с неоднородной толщиной b) Урановый куб в свинцовой оболочке}
    \end{center}
    \labelf{pic:sword}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=60mm]{figures/UranCube2.pdf} 
        \includegraphics[width=60mm]{figures/Hex.pdf}  
        \vspace{-3mm}
        \caption{a) Урановый куб в свинцовой оболочке b) Сравнение целлюлозы и RDX}
    \end{center}
    \labelf{pic:hex}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center} 
        \includegraphics[width=60mm]{figures/diffmat0.pdf} 
        \includegraphics[width=60mm]{figures/diffmat.pdf}  
        \vspace{-3mm}
        \caption{a) Энергетические спектры различных материалов (общий вид) 
        b) Энергетические спектры различных материалов (область с энергией более 4 МэВ)
    \end{center}
    \labelf{fig01}
    \vspace{-5mm}
\end{figure}
\begin{figure}[t]
    \begin{center}
        \includegraphics[width=60mm]{figures/diffmat1.pdf} 
        \includegraphics[width=60mm]{figures/Difference.pdf}  
        \vspace{-3mm}
        \caption{a) Зависимость метрики от эффективного заряда материала
 b) Сравнение энергетических спектров от уранового и алюминиевого шарика}
    \end{center}
    \labelf{pic:diff}
    \vspace{-5mm}
\end{figure}
На рисунке ~\ref{pic:sword})a пример опасного предмета из стали неравномерной толщины близкой к толщине стенок контейнера. На рисунках~\ref{pic:sword}b и ~\ref{pic:hex}a  показан результат  моделирования кубика урана с ребром 6 сантиметров (масса около 4 кг), помещённого в свинцовую сферу толщиной 1 см, как показало моделирования такой куб можно обнаружить при толщине оболочки вплоть до 5 сантиметров. Рисунок ~\ref{pic:hex}b демонстрирует различие между двумя органическими материалами: безопасным - целлюлозы, и опасным  - RDX. Различие значимо, а значит есть возможность создать алгоритмы поиска органических взрывчатых веществ. На рисуноке ~\ref{pic:diff}b  показывается результат сравнения двух энергетических спектров (в  качестве метрики выбран логарифм отношения интенсивностей) от алюминиевого и уранового шариков диаметром 1 см. Как мы видим даже на таких малых масштабах и небольших (по сравнение с реальным пучком) интенсивностях возможна регистрация различий в энергетических спектрах.

Также для проверки возможности определять эффективный заряд материала по энергетическому спектру было проведено
следующее моделирование. Было взято 6 мишений из различных материалов (), с одинаковым поперечным размером и
 различним продольным. Продольный размер подбирался таким образом, что бы полное ослабление потока гамма-лучей было
одинаковым для всех материалов, и их нельзя было различить только по анализу пространственого спектра.
Энегретические спектры этих мишений показаны на рисунках . Можно заметить все спектры различны в области до 3 Мэв
 (см. рис.) и в области после 4 Мэв (см. рис. ). Следует отметить что это отличие значимо даже для малых интесивностей
электронного  пучка ($10^8$), а значит на реальном электронном пучке с ускорителя ($10^{15}$) будет выражено более ярко.

\section*{Thickness reconstruction}
\begin{wrapfigure}[9]{r}{0.6\linewidth} 
    % Сделать подписи к отдельным картинкам
    \includegraphics[width=\linewidth]{figures/yed_schema_2.pdf}
    \vspace{-3mm}
    \caption{Thickness reconstruction}
    \labelf{schema2}
    \vspace{-5mm}
\end{wrapfigure}
Мы будем рассматривать одномерный случай, когда гамма-кванты проходят скозь стопку из нескольких материалов с фиксированной полной толщиной и нам нужно восстановить толщины отдельных материалов (см. схему ~\ref{scema2}). Мы будем использовать простую физическую модель в которой ослабление задается  следующей формулой: 
\begin{equation}
\label{eq:gamma}
\frac{N(E)}{N_0(E)} = \exp(-\sum_i \Sigma^{mean}_i(E)x_i),
\end{equation}
where $x_i$ --- thickness of the $i$-layer, $\Sigma^{mean}_i$ --- mean cross-section for group of materials with близкими зарядами ядра, $N,~N_0$ --- the number of gamma.При этом мы не будем учитывать множественное рассеяние и наличие аннигиляционной линии. Мы считаем что полная толщина известна и для восстановления толщин отдельных слоев мы будем использовать метод наименьших квадратов, то есть минимизировать такую сумму:
\begin{equation}
\sum_E(\ln \frac{N(E)}{N_0(E)} + \sum_i \Sigma^{mean}_i(E)x_i))^2 \to min
\end{equation}
Далее приведен пример работы алгоритма. We consider the object which has 3 layers of aluminium, iron and lead.The figure~\ref{rec:ex}a shows a contribution of every reconstructed material in summary attenuation. The table~\ref{tab:rec} contains the results of reconstruction.As we can see from the table, its result is quite accurate. To clarify the capabilities of the algorithm, we conducted several numerical experiments.We also used aluminum, iron and lead, but now we have taken many sets of thicknesses, with a total thickness between 30 and 180 centimeters. The figure~\ref{rec:ex}b shows the variation of recovery errors for these sets. As we can see, the thickness of heavy elements is determined best of all - with an accuracy about 5\%, and worst of all for materials from the iron group, the error there reaches 30\%. However, this is too simple model. What good will it do? We used only the energy distribution, if we also add a space distribution, then we can try to reconstruct the three-dimensional structure of the cargo in the container (the 3D gamma tomography). Therefore, our simple model shows that we have the prospect of creating a truly powerful system for analyzing the contents of containers.
\begin{figure}[t]
    \begin{center}
        % Сделать подписи к отдельным картинкам
        \includegraphics[width=60mm]{figures/reconstruction.pdf} 
        \includegraphics[width=60mm]{figures/relError.pdf}  
        \vspace{-3mm}
        \caption{a) Вклад отлельных слоев в общее ослабление потока b)Distribution of reconstruction errors for various numerical experiments}
    \end{center}
    \labelf{rec:ex}
    \vspace{-5mm}
\end{figure}

\begin{table}
    \caption{Пример результатов работы алгоритма востановления}
\begin{center}
        \begin{tabular}[c]{|c|c|c|}
        \hline 
        Material & Real, cm & Reconstructed, cm \\ 
        \hline 
        Al & 20 & 19.6 \\ 
        \hline 
        Fe & 40 & 41.6 \\ 
        \hline 
        Pb & 30 & 28.7 \\ 
        \hline 
    \end{tabular}
\end{center}
\caption{Example of thickness reconstruction}
    \label{tab:rec}
\end{table}

@@

The experiment was conducted by Dr. Guber and Dr. Ivashkin.
The sum of signals from 2 photodiode\\
Noise threshold: 100 KeV\\
\begin{table}
    \caption{Измерение энергетического разрешения детектора }
    \begin{center} 
        \begin{tabular}[c]{|c|c|c|}
